\input{header}

\begin{document}

\begin{center}
{\LARGE \textsc{Budget Justification}}
\end{center}
\hrule
\vspace{3mm}


\subsection*{Salaries and Wages Graduate PI}
The budget allocates summer salary for PI~Hirsch in all three years of the grant in the amount of \$15,053, \$15,505, and \$15,970, respectively. 
For the purpose of calculating the NSF two-month limit on salary for senior personnel, the University at Buffalo defines a ``year'' as its fiscal year, which is the consecutive 12-month period beginning July 1 and ending June 30.

\subsection*{Salaries and Wages Graduate Students}
Salaries and wages are based on the 2024 university budget with a 2\% annual increment.
The budget allocates salary funds for one graduate research assistant for all 3 years of the project. 
Per University policy, full-time graduate students may work a maximum of 20 hours per week during the academic year and up to 40 hours per week during academic breaks

\subsection*{Fringe Benefits}

Fringe benefit rates are based on the applicable federally negotiated rates.

\subsection*{Supplies}
The budget allocates \$2,000 in supplies in years 1 and 2 of the project, along with \$179 in supplies for year 3.
These will be used to purchase computational devices for use on this research project, including devices for students' use and experimental hardware.

\subsection*{Travel}
The budget allocates travel funds in each year of the project. 
The budget includes \$4,000 dollars per year for domestic travel.
This budget is for PI~Hirsch and one student to attend one domestic conference per year and to visit Madison once per year for collaboration.
The travel budget also includes \$6,000 for international travel for the PI and one student to attend one international conference per year.
Many of the specialized type systems conferences for choreographies (e.g. CONCUR) as well as the top programming languages either occur in Europe annually or rotate between the US and Europe.
Moreover, they are increasingly often hosted in Asia.
Because the location of conferences is not known more than a year in advance, the countries that we will travel to are currently unknown.

\subsection*{Tuition}
Tuition for graduate students is budgeted using the standard University at Buffalo projected tuition rates.

\subsection*{Computer Services Fee}
The budget requests funding for the University at Buffalo computer services required to complete the project.
These charges cover the cost of operating the CSE department's computing equipment, including servers, storage devices, workstations and printers used for research.
The amount assessed in each year takes into account the number of users of these facilities at a rate of \$... per person month of time on the project of CSE department personnel.
The computer services policy and formula are reviewed annually and charged uniformly to all sponsors. 

\subsection*{Indirect Costs}

The indirect costs are calculated at the University at Buffalo predetermined Facilities and Administrative (F\&A) cost rate of  60.1\% for all three years per DHHS agreement dated 09/04/2024.

\end{document}


%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: "akh-facilities"
%%% End:
