\input{header}

\begin{document}

\begin{center}
{\LARGE \textsc{Budget Justification}}
\end{center}
\hrule
\vspace{3mm}


\subsection*{Salaries and Wages Graduate PI}
The budget allocates summer salary for PI... in all ... years of the grant in the amount of ....
Note that PI... plan to consistently contribute to the project in all of its years, but the amount of his summer salary is reduced in the beginning of the project to comply with the NSF's rule of two months per year of NSF funding across all NSF-sponsored projects.
For the purpose of calculating the NSF two-month limit on salary for senior personnel, the University at Buffalo defines a ``year'' as its fiscal year, which is the consecutive 12-month period beginning July 1 and ending June 30.


\subsection*{Salaries and Wages Graduate Students}
Salaries and wages are based on the 2023 university budget with a 2\% annual increment.
The budget allocates salary funds for two graduate research assistants for all 3 years of the project. 
Per University policy, full-time graduate students may work a maximum of 20 hours per week during the academic year and up to 40 hours per week during academic breaks

\subsection*{Fringe Benefits}

Fringe benefit rates are based on the applicable federally negotiated rates.

\subsection*{Publication Costs}
The publication costs, budgeted at \$... in year 1 and in year 3, are intended to cover expenses associated with journal publishing, over-the-page-limit conference charges, and possible expenses associated with poster printing to present research results of the project.

\subsection*{Supplies}
The budget allocates \$... in supplies in year 1 of the project which will be used to ...

\subsection*{Travel}
The budget allocates travel funds in each year of the project. 
The budget includes \$... dollars per year for domestic travel.
This budget is for two people, either both students or the PI/co-PI and one student, to attend one domestic conference per year.
The travel budget also includes \$... for international travel for two people, either both students or the PI and one student, to attend one international conference per year.
Many of the specialized type systems conferences for session types (e.g. CONCUR) as well as the top programming languages either occur in Europe annually or rotate between the US and Europe.  
Because the location of conferences is not known more than a year in advance, the countries that we will travel to are currently unknown.

\subsection*{Tuition}
Tuition for graduate students is budgeted using the standard University at Buffalo projected tuition rates.

\subsection*{Computer Services Fee}
The budget requests funding for the University at Buffalo computer services required to complete the project.
These charges cover the cost of operating the CSE department's computing equipment, including servers, storage devices, workstations and printers used for research.
The amount assessed in each year takes into account the number of users of these facilities at a rate of \$... per person month of time on the project of CSE department personnel.
The computer services policy and formula are reviewed annually and charged uniformly to all sponsors. 

\subsection*{Indirect Costs}

The indirect costs are calculated at the University at Buffalo predetermined Facilities and Administrative (F\&A) cost rate of 60.5\% for Years 1 and 61\% for years 2-3 per DHHS agreement dated 04/27/2023.

\end{document}

