\input{header}


\begin{document}

\begin{center}
{\LARGE \textsc{Data Management Plan}}
\end{center}
\hrule

\vspace{3mm}

\subsection*{Types of Data}  
The data generated and produced under this grant will consist of proofs, language designs, and software as described in the proposal description.

\subsection*{Data Format and Content}
There is no set standard for the types of data expected to be generated under this proposal. However, PI Hirsch will use best practices for formatting and presenting the data generated and collected.
The code, files containing the formalism, and data generated will be made available on the PIs' personal websites as well as the project git repository.
The PI's websites are hosted by their respective universities and the project git repository is hosted on GitHub.

\subsection*{Facilities for Storage and Preservation}
All products of the proposed research program will be stored on departmental servers, where everything is regularly backed-up to a long-term archive.
Archives will be kept at University at Buffalo by the CSE IT staff as well as at the University of Wisconsin--Madison by CS IT staff.

\subsection*{Policies of Access} 
The proposed research program does not involve collecting sensitive data for evaluation; the proposed mechanisms are agnostic to the content of the data, {\em i.e.}, there is no need to look into the content of the data and analyze it.
Thus, all data gathered will be made publicly available.  

\subsection*{Policies for Reuse}  
Because the proposed research program does not involve collecting sensitive data for evaluation, data will be made available for reuse by other researchers.

\subsection*{Plans for Archiving} 
Data will be archived on the project website as well as with digital libraries that support data submission along with publications.
The PIs currently store data gathered in version-controlled repositories.
These repositories are synced with publication repositories.
This allows for easy correlation of which data was used for which publication.
In addition, data is organized by publication on the project website.
All data gathered, even if it is not used in a publication, is still associated with that publication and made available on the project website.  

\subsection*{Management Roles}
The PIs will be the primary managers of the data gathered and stored throughout the project.
After the grant period ends, the data collected will continue to be made available through the websites it was originally hosted on. 

\end{document}

