\input{header}

\usepackage{paralist}

\begin{document}

\begin{center}
{\LARGE
\textsc{Mentoring Plan}
}
\end{center}
\hrule

\vspace{3mm}

This grant provides for two graduate students,
one hosted at the University of Wisconsin--Madison and mentored by PI~Cecchetti,
and one hosted at the University at Buffalo and mentored by Co-PI~Hirsch.
These students will be mentored according to the plan below.

\paragraph{Setting and Communicating (Bidirectional) Expectations}

Each student will meet with their supervisor multiple times a week, including project meetings and one-on-one meetings.
Moreover, both PIs maintain an open-door policy: they makes time for impromptu questions and conversations with research students.
These policies ensure that there is a constant bidirectional line of communication between the senior personnel and their students.

Since the research is to be done collaboratively, we plan to have regular four-way communication.
The PIs have a multi-year history of a weekly zoom meeting including both PIs and students they both collaborate with.
These meetings are a priority, and are scheduled at the beginning of each semester with both PIs and both students
to ensure they continue to happen weekly.
The PIs also maintain a Slack instance where all collaborating students can ask questions and discuss concerns even outside of regular meeting times.

\paragraph{Individual Development Plan}

All students are different, and all students grow and mature during the course of their research experiences.
To support this, students will work with their supervisor to build an Individual Development Plan~(IDP) upon joining the group.
Work plans, timelines, and individual tasks will be set based on a combination of the project's needs and each graduate student's IDP.
They will be further refined and updated on a regular basis, informed in large part by the one-on-one meetings and open-door policy,
both of which allow a students to raise concerns and for the PI and co-PI to monitor their students' progress and needs (including needs they may not be able to detect or advocate for themselves).

Additionally, since skills, goals, and expectations change over time, the PI and co-PI will meet with students every semester to revise and update their IDPs.
This will provide a check-in point to encourage both students and supervisors to evaluate successes and failures, and make any need changes.


\paragraph{Promoting a Supportive Work Environment}

Both PIs believe strongly in lab culture, and work hard to promote participation in lab culture.
This provides students more than fun: it also provides a support network of students of all maturity levels and even faculty.
At the UW--Madison, PI~Cecchetti's group is an integral part of the programming languages research group (madPL),
including helping organize regular social events and building a strongly supportive and collaborative environment spanning the research groups of 5 faculty.
At UB, PI~Hirsch's  lab space is a significant source of graduate-student culture in the Department of Computer Science and Engineering.

Moreover, both PIs one-on-one meetings ensure that students have a time when they can bring up any issues which are concerning them.
This is one of the reasons that these one-on-one meetings are so important: they allow students to open up when they would not do so in other situations.

\paragraph{Assessing Progress}
Student progress is mostly measured by research progress and progress completing training tasks as described above.
Any issues they have are caught early thanks to the constant contact between the senior personnel and their students.
Moreover, both the UW--Madison and the Computer Science and Engineering department at UB require all students to be assessed annually.
Any students who raise concerns are then given mentoring by another department member.  Students are also evaluated annually at the department
level, which provides an external evaluation of progress.

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
