\input{header}

\begin{document}

\section*{Project Summary}

Choreographic programming is a new programming paradigm promising to simplify development of concurrent and distributed systems.
Unlike traditional paradigms where developers write one program for each system component (e.g., node, process, or thread),
a choreography consists of a single top-level program that specifies the behaviors of each component and how they interact.
This is a powerful new perspective, but existing choreographies still lack the flexibility to express many common patterns in concurrent applications.

A recent result by the PIs lays new groundwork to address these gaps.
Choreographies can now specify programs that dynamically compute and communicate which party should perform a given task---%%%
a feature known as first-class process polymorphism.
This proposal \emph{leverages first-class process polymorphism to design and implement crucial new features and develop the theory surrounding them},
giving choreographies the ability to describe a much larger swath of systems and bringing them closer to their promise.

\subsection*{Intellectual Merit}

To expand the expressive power of choreographic programming, this project will
(1)~develop a suite of important new features within the choreographic model,
including \emph{dynamic thread spawning and killing}
and threads waiting to receive a set of messages satisfying \emph{developer-specified requirements} before continuing,
that are made possible by the machinery of first-class process polymorphism;
(2)~\emph{eliminate substantial performance bottlenecks} associated with current implementations of process polymorphism
and build a \emph{representative benchmark suite} to enable more rigorous performance analysis by the entire choreographic programming community;
and (3)~develop new fundamental theory to enable \emph{choreographic services},
processes that receive and process messages from arbitrary clients without knowing a client's identity before a request arrives.

The outcomes of this proposal will include novel choreographic features and theory needed to prove
important correctness properties about those features, improved performance of choreographic implementations,
and a public benchmark suite for compilation of process polymorphism.
The PIs bring together a proven track record of fruitful collaboration
and substantial expertise in choreographies, programming-language theory, and applications of the same to systems.

\subsection*{Broader Impacts}

If successful, this research will significantly expand the expressive power of choreographic programming.
It will remove a core barrier impeding the use of choreographic techniques for practical concurrent and distributed systems
and create a foundation for future research in both programming languages and concurrent programs.
It will also produce improvements to nascent choreographic compiler technology
and the first public benchmark suite specifically to measure choreographic compiler performance.

Creating more practical choreographic technology, including concrete implementation work,
will enable teaching undergraduate and graduate students about this emerging new paradigm
and create research opportunities for undergraduate students.
PI~Cecchetti will incorporate choreographic theories into his ``Principles of Programming Languages'' class at UW--Madison,
and PI~Hirsch will involve undergraduates in the implementation components of this research as part of the University at Buffalo's ``Experiential Learning and Research'' course.
This course allows undergraduate students to work with faculty in small teams over three semesters on a project.
As part of this program, PI~Hirsch teaches undergraduates about choreographic programming, concurrency, compilers, and functional programming.

\end{document}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
